---
title: "Publication-Ready Tables in R"
subtitle: "gtsummary — from basics to customisation with flextable"
author: "Workshop"
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    scrollable: true
execute:
  echo: true
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(gtsummary)
library(flextable)
library(janitor)

# We will use the built-in trial dataset from gtsummary
# It mimics a clinical trial with treatment arms, demographics, and outcomes
data(trial)
glimpse(trial)
```

---

# Part 1 — gtsummary Basics {background-color="#2C3E50"}

## What is gtsummary?

- `gtsummary` creates **publication-ready summary tables** with minimal code
- Automatically detects variable type (continuous vs categorical)
- Adds statistics, p-values, and confidence intervals
- Outputs are compatible with Word, PDF, HTML, and PowerPoint

```{r}
# Install if needed
# install.packages("gtsummary")
library(gtsummary)
```

---

## tbl_summary() — Simplest Table

```{r}
trial |>
  select(age, grade, response, trt) |>
  tbl_summary()
```

*`tbl_summary()` auto-detects variable types and shows median (IQR) for continuous, n (%) for categorical.*

---

## tbl_summary() — By Group

Stratify columns by a grouping variable using `by =`.

```{r}
trial |>
  select(age, grade, response, trt) |>
  tbl_summary(by = trt)
```

---

## Adding Overall Column

```{r}
trial |>
  select(age, grade, response, trt) |>
  tbl_summary(by = trt) |>
  add_overall()
```

---

## Adding P-values

```{r}
trial |>
  select(age, grade, response, trt) |>
  tbl_summary(by = trt) |>
  add_overall() |>
  add_p()
```

> `add_p()` automatically selects the appropriate test: Wilcoxon rank-sum for continuous, Chi-square/Fisher for categorical.

---

## Customising Statistics

Change what statistics are shown for each variable type.

```{r}
trial |>
  select(age, marker, grade, trt) |>
  tbl_summary(
    by = trt,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    )
  )
```

> Use `{mean}`, `{sd}`, `{median}`, `{p25}`, `{p75}`, `{n}`, `{p}`, `{N}` as placeholders.

---

## Formatting Digits

```{r}
trial |>
  select(age, marker, trt) |>
  tbl_summary(
    by = trt,
    statistic = all_continuous() ~ "{mean} ({sd})",
    digits = all_continuous() ~ 1
  )
```

---

## Custom Labels

```{r}
trial |>
  select(age, grade, trt) |>
  tbl_summary(
    by = trt,
    label = list(
      age   ~ "Age (years)",
      grade ~ "Tumour Grade"
    )
  )
```

---

## Changing Missing Data Label

```{r}
trial |>
  select(age, grade, trt) |>
  tbl_summary(
    by = trt,
    missing_text = "Not recorded"
  )
```

---

## Including / Excluding Missing

```{r}
# Show no missing row at all
trial |>
  select(age, grade, trt) |>
  tbl_summary(
    by  = trt,
    missing = "no"
  )
```

---

## tbl_summary() — All Together

```{r}
trial |>
  select(age, marker, grade, response, trt) |>
  tbl_summary(
    by = trt,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits  = all_continuous() ~ 1,
    label   = list(
      age      ~ "Age (years)",
      marker   ~ "Marker level",
      grade    ~ "Tumour Grade",
      response ~ "Tumour Response"
    ),
    missing_text = "Not recorded"
  ) |>
  add_overall() |>
  add_p()
```

---

# Part 2 — Adding More Information {background-color="#2C3E50"}

## add_n() — Sample Sizes per Column

```{r}
trial |>
  select(age, grade, trt) |>
  tbl_summary(by = trt) |>
  add_n()
```

---

## add_stat_label() — Describe the Statistic Used

```{r}
trial |>
  select(age, grade, trt) |>
  tbl_summary(by = trt) |>
  add_stat_label()
```

---

## bold_labels() and italicize_levels()

```{r}
trial |>
  select(age, grade, trt) |>
  tbl_summary(by = trt) |>
  bold_labels() |>
  italicize_levels()
```

---

## bold_p() — Highlight Significant P-values

```{r}
trial |>
  select(age, grade, response, trt) |>
  tbl_summary(by = trt) |>
  add_p() |>
  bold_p(t = 0.05)
```

---

## sort_p() — Sort by P-value

```{r}
trial |>
  select(age, grade, response, trt) |>
  tbl_summary(by = trt) |>
  add_p() |>
  sort_p()
```

---

# Part 3 — Regression Tables {background-color="#2C3E50"}

## tbl_regression() — Logistic Regression

```{r}
m_logistic <- glm(response ~ age + grade + trt,
                  data   = trial,
                  family = binomial)

tbl_regression(m_logistic, exponentiate = TRUE)
```

> `exponentiate = TRUE` shows **Odds Ratios** instead of log-odds.

---

## tbl_regression() — Linear Regression

```{r}
m_linear <- lm(marker ~ age + grade + trt, data = trial)

tbl_regression(m_linear)
```

---

## add_glance_table() — Model Fit Statistics

```{r}
m_logistic |>
  tbl_regression(exponentiate = TRUE) |>
  add_glance_table(include = c(nobs, AIC, BIC))
```

---

## tbl_uvregression() — Univariable Table

Run univariable regressions for each predictor against one outcome in one step.

```{r}
trial |>
  select(age, grade, marker, trt, response) |>
  tbl_uvregression(
    method    = glm,
    y         = response,
    method.args = list(family = binomial),
    exponentiate = TRUE
  ) |>
  add_global_p()
```

---

# Part 4 — Merging and Stacking {background-color="#2C3E50"}

## tbl_merge() — Side by Side Tables

Combine univariable and multivariable regression tables.

```{r}
tbl_uv <- trial |>
  select(age, grade, marker, trt, response) |>
  tbl_uvregression(
    method       = glm,
    y            = response,
    method.args  = list(family = binomial),
    exponentiate = TRUE
  )

tbl_mv <- glm(response ~ age + grade + marker + trt,
              data   = trial,
              family = binomial) |>
  tbl_regression(exponentiate = TRUE)

tbl_merge(
  tbls        = list(tbl_uv, tbl_mv),
  tab_spanner = c("**Univariable**", "**Multivariable**")
)
```

---

## tbl_stack() — Stack Tables Vertically

```{r}
tbl1 <- trial |>
  filter(trt == "Drug A") |>
  select(age, grade, response) |>
  tbl_summary() |>
  modify_header(stat_0 ~ "**Drug A** (n = {n})")

tbl2 <- trial |>
  filter(trt == "Drug B") |>
  select(age, grade, response) |>
  tbl_summary() |>
  modify_header(stat_0 ~ "**Drug B** (n = {n})")

tbl_stack(list(tbl1, tbl2), group_header = c("Drug A", "Drug B"))
```

---

# Part 5 — Table Modifications {background-color="#2C3E50"}

## modify_header() — Custom Column Headers

```{r}
trial |>
  select(age, grade, trt) |>
  tbl_summary(by = trt) |>
  modify_header(
    label       ~ "**Characteristic**",
    stat_1      ~ "**Drug A** (n = {n})",
    stat_2      ~ "**Drug B** (n = {n})"
  )
```

---

## modify_caption() — Add a Table Caption

```{r}
trial |>
  select(age, grade, trt) |>
  tbl_summary(by = trt) |>
  modify_caption("**Table 1. Baseline Characteristics by Treatment Group**")
```

---

## modify_footnote() — Add Footnotes

```{r}
trial |>
  select(age, grade, trt) |>
  tbl_summary(by = trt) |>
  add_p() |>
  modify_footnote(
    p.value ~ "Wilcoxon rank-sum test; Chi-square test"
  )
```

---

## modify_spanning_header() — Group Columns

```{r}
trial |>
  select(age, grade, trt) |>
  tbl_summary(by = trt) |>
  modify_spanning_header(
    c(stat_1, stat_2) ~ "**Treatment Group**"
  )
```

---

## Removing Rows with modify_table_body()

```{r}
trial |>
  select(age, grade, trt) |>
  tbl_summary(by = trt) |>
  modify_table_body(
    ~ .x |> filter(!(variable == "grade" & row_type == "missing"))
  )
```

---

# Part 6 — Exporting gtsummary Tables {background-color="#2C3E50"}

## Converting to Different Formats

```{r eval=FALSE}
tbl <- trial |>
  select(age, grade, response, trt) |>
  tbl_summary(by = trt) |>
  add_p()

# To Word (.docx) — requires officer package
tbl |> as_flex_table() |> save_as_docx(path = "table1.docx")

# To HTML
tbl |> as_gt() |> gt::gtsave("table1.html")

# To PDF (via gt)
tbl |> as_gt() |> gt::gtsave("table1.pdf")
```

---

# Part 7 — Modifying with flextable {background-color="#2C3E50"}

## What is flextable?

- `flextable` provides **fine-grained control** over table appearance
- Works seamlessly with `gtsummary` via `as_flex_table()`
- Ideal for Word/PowerPoint output
- Controls: fonts, cell colours, borders, column widths, alignment, footnotes

```{r}
library(flextable)
```

---

## Converting gtsummary to flextable

```{r}
tbl <- trial |>
  select(age, grade, response, trt) |>
  tbl_summary(
    by      = trt,
    missing = "no"
  ) |>
  add_overall() |>
  add_p() |>
  bold_labels()

ft <- as_flex_table(tbl)
ft
```

---

## autofit() — Auto Column Widths

```{r}
ft |> autofit()
```

> Always run `autofit()` first — it resizes columns to fit content before applying other formatting.

---

## set_table_properties() — Table Width

```{r}
ft |>
  autofit() |>
  set_table_properties(width = 1, layout = "autofit")
```

> `width = 1` means full page width. Values between 0 and 1 set proportional width.

---

## theme_vanilla() and Other Themes

```{r}
ft |> theme_vanilla()
ft |> theme_box()
ft |> theme_booktabs()
ft |> theme_alafoli()
ft |> theme_zebra()
```

---

## Bold Header

```{r}
ft |>
  autofit() |>
  bold(part = "header")
```

---

## Font Face and Size

```{r}
ft |>
  autofit() |>
  font(fontname = "Times New Roman", part = "all") |>
  fontsize(size = 11, part = "all") |>
  fontsize(size = 12, part = "header")
```

---

## Aligning Columns

```{r}
ft |>
  autofit() |>
  align(align = "center", part = "header") |>
  align(j    = 2:5, align = "center", part = "body")
```

> `j` selects columns by position. `part` can be `"header"`, `"body"`, or `"all"`.

---

## Colouring Cells — bg()

Highlight specific cells by value or condition.

```{r}
ft |>
  autofit() |>
  bg(part = "header", bg = "#2C3E50") |>
  color(part = "header", color = "white")
```

---

## Conditional Highlighting — p-value Column

```{r}
# Highlight rows where p < 0.05 in light yellow
ft |>
  autofit() |>
  bg(
    j = "p.value",
    bg = "lightyellow",
    part = "body"
  )
```

---

## Adding Borders — hline() and vline()

```{r}
border_style <- officer::fp_border(color = "black", width = 1.5)

ft |>
  autofit() |>
  hline_top(part = "header", border = border_style) |>
  hline_bottom(part = "header", border = border_style) |>
  hline_bottom(part = "body", border = border_style)
```

---

## Merging Header Cells — merge_h()

```{r}
ft |>
  autofit() |>
  add_header_row(
    values = c("", "Treatment Arms", ""),
    colwidths = c(1, 2, 2)
  ) |>
  merge_h(part = "header") |>
  bold(part = "header") |>
  align(align = "center", part = "header")
```

---

## set_caption() — Table Caption

```{r}
ft |>
  autofit() |>
  set_caption(
    caption = "Table 1. Baseline characteristics by treatment group",
    style   = "Table Caption"
  )
```

---

## add_footer_lines() — Footnotes

```{r}
ft |>
  autofit() |>
  add_footer_lines(
    "IQR: Interquartile Range. p-values: Wilcoxon rank-sum (continuous), Chi-square (categorical)."
  ) |>
  italic(part = "footer")
```

---

## Padding Cells — padding()

```{r}
ft |>
  autofit() |>
  padding(padding = 4, part = "all")
```

---

## Complete Formatted Table

```{r}
border_h <- officer::fp_border(color = "#2C3E50", width = 1.5)
border_l <- officer::fp_border(color = "#cccccc", width = 0.5)

ft |>
  autofit() |>
  theme_booktabs() |>
  bold(part = "header") |>
  font(fontname = "Calibri", part = "all") |>
  fontsize(size = 10, part = "body") |>
  fontsize(size = 11, part = "header") |>
  bg(part = "header", bg = "#2C3E50") |>
  color(part = "header", color = "white") |>
  align(align = "center", part = "header") |>
  align(j = 2:5, align = "center", part = "body") |>
  hline_top(part = "header", border = border_h) |>
  hline_bottom(part = "body", border = border_h) |>
  padding(padding = 5, part = "all") |>
  add_footer_lines(
    "IQR: Interquartile Range. Chi-square or Wilcoxon tests used."
  ) |>
  italic(part = "footer") |>
  set_caption("Table 1. Baseline Characteristics by Treatment Group")
```

---

## Saving to Word

```{r eval=FALSE}
ft |>
  autofit() |>
  theme_booktabs() |>
  save_as_docx(path = "table1_formatted.docx")
```

---

## Saving to PowerPoint

```{r eval=FALSE}
ft |>
  autofit() |>
  save_as_pptx(path = "table1_formatted.pptx")
```

---

## Quick Reference — Key Functions

| Task | gtsummary | flextable |
|---|---|---|
| Basic table | `tbl_summary()` | `as_flex_table()` |
| By group | `by =` | — |
| P-values | `add_p()` | — |
| Bold labels | `bold_labels()` | `bold(part = "header")` |
| Caption | `modify_caption()` | `set_caption()` |
| Footnote | `modify_footnote()` | `add_footer_lines()` |
| Font | — | `font()`, `fontsize()` |
| Cell colour | — | `bg()`, `color()` |
| Borders | — | `hline_top()`, `hline_bottom()` |
| Alignment | — | `align()` |
| Auto width | — | `autofit()` |
| Export Word | `as_flex_table() + save_as_docx()` | `save_as_docx()` |
