---
title: "Tidyverse Data Wrangling Practice"
subtitle: "Slide Exercises"
author: "Sparrow"
date: today
format: 
  revealjs:
    theme: serif
    css: ../assets/theme_webp.css
    include-after-body: footer.html
    slide-number: true
    navigation-mode: linear
    code-copy: true
    code-fold: true
    code-summary: "Show the code"
    chalkboard: true
    preview-links: true
    center: true
    zoom: true
    echo: true
---

## How to Use

- Each slide contains one tidyverse practice question.
- Fill in the blanks marked as `_____`.
- These tasks cover core verbs, recoding, reshaping, joins, and cleaning.

## Setup Data

```{r}
library(tidyverse)

clinic_wide <- tibble(
  id = 1:6,
  sex = c("Female", "Male", "Female", "Male", "Female", "Male"),
  age = c(22, 34, 29, 41, 37, 26),
  bmi = c(19.8, 27.2, 31.1, 24.3, 29.4, 22.6),
  fbs_0 = c(88, 110, 132, 96, 124, 91),
  fbs_3 = c(86, 104, 120, 94, 118, 89),
  ppbs_0 = c(126, 168, 214, 142, 176, 130),
  smoker = c("No", "Yes", "No", "Yes", "No", "No")
)

clinic_long <- tibble(
  id = c(1,1,2,2,3,3,4,4),
  visit = c("baseline", "3m", "baseline", "3m", "baseline", "3m", "baseline", "3m"),
  fbs = c(88, 86, 110, 104, 132, 120, 96, 94)
)

treatment_map <- tibble(
  id = 1:6,
  treatment = c("Diet", "Metformin", "Insulin", "Diet", "Metformin", "Diet")
)
```

## Exercise 1

Select only `id`, `age`, and `bmi`.

```{r eval=FALSE}
clinic_wide %>%
  _____(id, age, bmi)
```

## Exercise 2

Filter participants with BMI 25 or more.

```{r eval=FALSE}
clinic_wide %>%
  _____(bmi >= 25)
```

## Exercise 3

Create a new column `bmi_group` where BMI is classified as:

- `<25` = `"Normal"`
- `25 to <30` = `"Overweight"`
- `>=30` = `"Obese"`

```{r eval=FALSE}
clinic_wide %>%
  mutate(
    bmi_group = case_when(
      bmi < 25 ~ "Normal",
      bmi >= 25 & bmi < 30 ~ "Overweight",
      bmi >= 30 ~ "Obese"
    )
  )
```

## Exercise 4

Create `age_group` as `"Under30"` or `"30plus"` using `mutate()`.

```{r eval=FALSE}
clinic_wide %>%
  _____(age_group = if_else(age < 30, "Under30", "30plus"))
```

## Exercise 5

Arrange rows from highest to lowest baseline FBS.

```{r eval=FALSE}
clinic_wide %>%
  _____(desc(fbs_0))
```

## Exercise 6

Find mean BMI by sex.

```{r eval=FALSE}
clinic_wide %>%
  _____(sex) %>%
  summarise(mean_bmi = mean(bmi))
```

## Exercise 7

Count participants by smoking status.

```{r eval=FALSE}
clinic_wide %>%
  _____(smoker)
```

## Exercise 8

Convert `clinic_wide` from wide to long for the two FBS columns.

```{r eval=FALSE}
clinic_wide %>%
  pivot_longer(
    cols = _____,
    names_to = "visit",
    values_to = "fbs"
  )
```

## Exercise 9

Convert `clinic_long` from long to wide so each visit becomes a separate column.

```{r eval=FALSE}
clinic_long %>%
  pivot_wider(
    names_from = _____,
    values_from = fbs
  )
```

## Exercise 10

Join treatment information into `clinic_wide` using `id`.

```{r eval=FALSE}
clinic_wide %>%
  _____(treatment_map, by = "id")
```

## Exercise 11

Rename `fbs_0` to `fbs_baseline`.

```{r eval=FALSE}
clinic_wide %>%
  _____(fbs_baseline = fbs_0)
```

## Exercise 12

Move `smoker` to appear right after `sex`.

```{r eval=FALSE}
clinic_wide %>%
  _____(smoker, .after = sex)
```

## Exercise 13

Keep only distinct combinations of `sex` and `smoker`.

```{r eval=FALSE}
clinic_wide %>%
  _____(sex, smoker)
```

## Exercise 14

Create a flag `high_fbs` that is `TRUE` when baseline FBS is 126 or more.

```{r eval=FALSE}
clinic_wide %>%
  mutate(high_fbs = _____ >= 126)
```

## Exercise 15

Create a variable `fbs_change` equal to 3-month FBS minus baseline FBS.

```{r eval=FALSE}
clinic_wide %>%
  mutate(fbs_change = _____ - fbs_0)
```

## Exercise 16

Use `select()` to keep all columns except `ppbs_0`.

```{r eval=FALSE}
clinic_wide %>%
  select(-_____)
```

## Exercise 17

Filter female participants younger than 30 years.

```{r eval=FALSE}
clinic_wide %>%
  filter(sex == "Female", _____ < 30)
```

## Exercise 18

Summarise the maximum baseline PPBS.

```{r eval=FALSE}
clinic_wide %>%
  summarise(max_ppbs = _____(ppbs_0))
```

## Exercise 19

Create a categorical FBS variable from baseline values:

- `<100` = `"Normal"`
- `100 to 125` = `"Prediabetes"`
- `>=126` = `"Diabetes"`

```{r eval=FALSE}
clinic_wide %>%
  mutate(
    fbs_status = case_when(
      fbs_0 < 100 ~ "Normal",
      fbs_0 >= 100 & fbs_0 < 126 ~ "Prediabetes",
      fbs_0 >= 126 ~ "Diabetes"
    )
  )
```

## Exercise 20

Group by sex and summarise mean baseline FBS and mean baseline PPBS.

```{r eval=FALSE}
clinic_wide %>%
  group_by(sex) %>%
  summarise(
    mean_fbs = mean(_____),
    mean_ppbs = mean(ppbs_0)
  )
```

## Instructor Key

```{r eval=FALSE}
# 1 select
# 2 filter
# 4 mutate
# 5 arrange
# 6 group_by
# 7 count
# 8 c(fbs_0, fbs_3)
# 9 visit
#10 left_join
#11 rename
#12 relocate
#13 distinct
#14 fbs_0
#15 fbs_3
#16 ppbs_0
#17 age
#18 max
#20 fbs_0
```
