---
title: "Advanced Tidyverse Wrangling Practice"
subtitle: "Task-Based Quarto Worksheet"
author: "Workshop Practice"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This worksheet focuses on more applied data wrangling tasks using tidyverse. The dataset is small, but the tasks reflect real workflows used in health and clinical research projects.

## Packages

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Data

```{r}
patients <- tibble(
  patient_id = 1001:1010,
  sex = c("Female", "Male", "Female", "Male", "Female", "Male", "Female", "Male", "Female", "Male"),
  age = c(24, 46, 31, 52, 28, 37, 41, 33, 26, 49),
  district = c("Bareilly", "Lucknow", "Bareilly", "Agra", "Lucknow", "Bareilly", "Agra", "Lucknow", "Bareilly", "Agra"),
  visit_date = c("2026-01-05", "2026-01-07", "2026-01-09", "2026-01-11", "2026-01-14", "2026-01-15", "2026-01-20", "2026-01-21", "2026-01-23", "2026-01-25"),
  bmi = c(20.1, 27.8, 24.2, 31.4, 26.0, 25.2, 29.1, 23.7, 21.8, 32.5),
  treatment = c("Diet", "Metformin", "Diet", "Insulin", "Metformin", "Diet", "Metformin", "Diet", "Diet", "Insulin")
)

glucose_wide <- tibble(
  patient_id = 1001:1010,
  fbs_0 = c(86, 112, 94, 138, 101, 109, 126, 98, 92, 144),
  fbs_3 = c(84, 104, 91, 128, 97, 102, 118, 95, 89, 136),
  ppbs_0 = c(118, 166, 132, 214, 156, 172, 201, 149, 128, 226),
  ppbs_3 = c(112, 154, 126, 198, 148, 160, 186, 143, 121, 210)
)

hba1c <- tibble(
  patient_id = 1001:1010,
  hba1c_baseline = c(5.3, 6.4, 5.8, 8.1, 6.0, 6.2, 7.1, 5.9, 5.5, 8.5),
  hba1c_3m = c(5.2, 6.0, 5.6, 7.3, 5.7, 5.9, 6.6, 5.8, 5.3, 7.8)
)
```

## Task 1: Clean and inspect

Clean the data and prepare it for analysis.

- Convert `visit_date` to a date.
- Apply `clean_names()`.
- Check structure and preview rows.

```{r}
# Write your code here
```

## Task 2: Join files

Join `patients`, `glucose_wide`, and `hba1c` into one analysis dataset called `analysis_df` using `patient_id`.

```{r}
# Write your code here
```

## Task 3: Continuous to categorical

Using `mutate()` and `case_when()`, create the following variables in `analysis_df`.

- `age_group`: `<30`, `30-39`, `40+`
- `bmi_group`: `Normal`, `Overweight`, `Obese`
- `fbs_status`: `Normal`, `Prediabetes`, `Diabetes` based on `fbs_0`

```{r}
# Write your code here
```

## Task 4: Change scores

Create these new variables.

- `fbs_change = fbs_3 - fbs_0`
- `ppbs_change = ppbs_3 - ppbs_0`
- `hba1c_change = hba1c_3m - hba1c_baseline`

Then create a flag called `improved_hba1c` for values below 0.

```{r}
# Write your code here
```

## Task 5: Wide to long pivot

Convert the repeated glucose columns into a long dataset named `glucose_long`.

Expected idea:
- Keep `patient_id`
- Gather `fbs_0`, `fbs_3`, `ppbs_0`, `ppbs_3`
- Separate test type and time point

```{r}
# Write your code here
```

## Task 6: Long to wide pivot

From `glucose_long`, create a tidy wide table where each row is one patient-time combination and glucose types become separate columns.

Suggested output columns:
- `patient_id`
- `time`
- `fbs`
- `ppbs`

```{r}
# Write your code here
```

## Task 7: District summary

Create a district-wise summary with:

- `n`
- `mean_age`
- `mean_bmi`
- `mean_fbs_0`
- `mean_ppbs_0`
- `mean_hba1c_change`

Arrange from highest to lowest `mean_fbs_0`.

```{r}
# Write your code here
```

## Task 8: High-risk subset

Create `high_risk_df` for participants meeting any of these:

- `fbs_0 >= 126`
- `ppbs_0 >= 200`
- `bmi >= 30`

Keep only the most relevant review columns.

```{r}
# Write your code here
```

## Task 9: Treatment-level report

For each treatment group, calculate:

- total participants
- diabetes count
- obesity count
- mean baseline FBS
- mean HbA1c change

```{r}
# Write your code here
```

## Task 10: Grouped clinical summary

Create a summary by `sex` and `age_group` showing:

- `n`
- `mean_bmi`
- `mean_fbs_0`
- `mean_ppbs_0`

Round numeric summaries to 1 decimal place.

```{r}
# Write your code here
```

## Task 11: Task-specific filter pipeline

Build a single pipeline that returns only Bareilly participants who are either prediabetic or diabetic. Select a compact review table including demographics, treatment, and glucose values.

```{r}
# Write your code here
```

## Task 12: Dates and ordering

Using the visit date:

- Extract month and week number.
- Arrange from earliest to latest visit.
- Create a dataset of the first visit in each district.

```{r}
# Write your code here
```

## Task 13: Missing-data style practice

Create a modified copy of `analysis_df` where you intentionally replace a few values with `NA`, then:

- count missing values per column,
- drop rows with missing `fbs_0`,
- replace missing `treatment` with `"Unknown"`.

```{r}
# Write your code here
```

## Task 14: Export-ready final summary

Create `final_summary` with one row per `district` and `treatment` containing:

- `n`
- `median_fbs_0`
- `median_ppbs_0`
- `mean_bmi`
- `improved_hba1c_n`

```{r}
# Write your code here
```

## Optional extension

Try any of these extra tasks.

- Write `final_summary` to CSV.
- Plot mean baseline FBS by district.
- Identify the top 3 participants with the largest HbA1c reduction.
- Reorder `bmi_group` as an ordered factor.
- Use `relocate()` to make derived variables appear together.
