---
title: "ggstatsplot — Intermediate"
format:
  html:
    code-copy: true
    toc: true
    lightbox: auto
---

::: {.callout-note}
This module assumes familiarity with the basics covered in
[ggstatsplot — Introduction](ggstatsplot_intro.qmd).
:::

## Functions covered in this module

| Function | Use case |
|---|---|
| `ggwithinstats()` | Repeated measures / paired comparisons |
| `ggbarstats()` | Proportions across categorical groups |
| `ggpiestats()` | Proportions as pie / donut charts |
| Customising appearance | Colours, themes, subtitle text |
| Interpreting Bayes factors | Introduction |

---

## `ggwithinstats()` — Repeated Measures

Use this when the same participants are measured more than once (paired/within-subjects design) — for example, pre-post measurements.

```{r}
#| eval: false
library(ggstatsplot)
library(tidyr)

# Reshape from wide to long for within-subjects comparison
overall_long <- overall_up |>
  select(member_id, sbp_0m, sbp_3m) |>
  pivot_longer(
    cols      = c(sbp_0m, sbp_3m),
    names_to  = "timepoint",
    values_to = "sbp"
  ) |>
  mutate(timepoint = factor(timepoint,
                            levels = c("sbp_0m", "sbp_3m"),
                            labels = c("Baseline", "3 Months")))

ggwithinstats(
  data    = overall_long,
  x       = timepoint,
  y       = sbp,
  subject.id = member_id,       # links pairs across timepoints
  title   = "Systolic BP: Baseline vs 3 Months"
)
```

**Key differences from `ggbetweenstats()`:**

- `subject.id` is required — it tells the function which rows are paired
- Uses a paired t-test (parametric) or Wilcoxon signed-rank (non-parametric)
- Lines connect paired observations in the plot

---

## `ggbarstats()` — Proportions

Use this when you want to compare proportions of a categorical outcome across groups. The output includes a χ² or Fisher's exact test.

```{r}
#| eval: false
ggbarstats(
  data  = overall_up,
  x     = fuel_type,           # grouping variable (x-axis)
  y     = tobacco_type,        # outcome category (fill)
  title = "Tobacco Type by Fuel Type"
)
```

**What the output shows:**

- Stacked bar chart with proportion labels
- χ² test statistic, degrees of freedom, p-value, and Cramér's V effect size
- Pairwise comparisons between bars (if > 2 groups)

**Swap axes to change orientation:**

```{r}
#| eval: false
# Make tobacco_type the grouping variable
ggbarstats(data = overall_up, x = tobacco_type, y = fuel_type)
```

---

## `ggpiestats()` — Pie / Donut Charts

Use this for a single categorical variable, or two categorical variables where a pie chart is appropriate for the audience.

```{r}
#| eval: false
# Single variable — proportions of intervention types
ggpiestats(
  data  = overall_up,
  x     = intervention,
  title = "Distribution of Intervention Groups"
)
```

```{r}
#| eval: false
# Two variables — tobacco type by intervention
ggpiestats(
  data  = overall_up,
  x     = tobacco_type,
  y     = intervention,       # creates faceted pies
  title = "Tobacco Type by Intervention"
)
```

::: {.callout-warning}
## When to use pie charts
Pie charts are appropriate when there are ≤ 5 categories and the audience
is non-technical. For more categories or detailed comparisons, prefer
`ggbarstats()`.
:::

---

## Customising appearance

All ggstatsplot functions return a ggplot2 object, so you can extend them normally.

```{r}
#| eval: false
library(ggplot2)

ggbetweenstats(
  data        = overall_up,
  x           = fuel_type,
  y           = sbp_0m,
  # Colours for each group
  palette     = "Set2",
  package     = "RColorBrewer",
  # Remove the subtitle (statistical results) if preferred
  results.subtitle = TRUE,
  # Change point size and transparency
  point.args  = list(size = 2, alpha = 0.4),
  title       = "Systolic BP by Fuel Type"
) +
  theme_minimal(base_size = 13) +
  labs(caption = "Source: NCTOH Study Dataset")
```

**Common customisation parameters:**

| Parameter | What it controls |
|---|---|
| `palette` + `package` | Colour palette for groups |
| `results.subtitle` | Show/hide statistical annotation |
| `point.args` | Size, alpha, shape of data points |
| `centrality.plotting` | Show/hide mean/median marker |
| `xlab` / `ylab` | Axis labels |

---

## Introduction to Bayes Factors

When you set `type = "bayes"`, ggstatsplot reports a **Bayes Factor (BF₁₀)** instead of a p-value.

```{r}
#| eval: false
ggbetweenstats(
  data  = overall_up,
  x     = fuel_type,
  y     = sbp_0m,
  type  = "bayes",
  title = "Systolic BP by Fuel Type (Bayesian)"
)
```

**Interpreting BF₁₀:**

| BF₁₀ | Interpretation |
|---|---|
| > 100 | Extreme evidence for H₁ |
| 30–100 | Very strong evidence for H₁ |
| 10–30 | Strong evidence for H₁ |
| 3–10 | Moderate evidence for H₁ |
| 1–3 | Anecdotal evidence for H₁ |
| ~1 | No evidence either way |
| 1/3–1 | Anecdotal evidence for H₀ |
| < 1/10 | Strong–extreme evidence for H₀ |

- **BF₁₀ > 3**: reasonably confident that H₁ (an effect exists) is supported
- **BF₁₀ < 1/3**: reasonably confident that H₀ (no effect) is supported
- Values between 1/3 and 3 are inconclusive

::: {.callout-note}
Bayesian inference and deeper interpretation of Bayes factors is covered in
[ggstatsplot — Advanced](ggstatsplot_advanced.qmd).
:::

---

## Quick reference

```{r}
#| eval: false
# Paired/repeated measures
ggwithinstats(data = df, x = timepoint, y = outcome, subject.id = id)

# Proportions (bar)
ggbarstats(data = df, x = group, y = category)

# Proportions (pie)
ggpiestats(data = df, x = category)

# Bayesian version of any function
ggbetweenstats(data = df, x = group, y = outcome, type = "bayes")
```
