---
title: "ggstatsplot — Advanced"
format:
  html:
    code-copy: true
    toc: true
    lightbox: auto
---

::: {.callout-note}
This module builds on [ggstatsplot — Introduction](ggstatsplot_intro.qmd) and
[ggstatsplot — Intermediate](ggstatsplot_moderate.qmd).
:::

## Topics covered

- Full Bayesian reporting workflow
- `grouped_*` functions — faceted comparisons
- `ggcorrmat()` — correlation matrices
- Combining with `patchwork`
- Publication-ready output
- Healthcare case study

---

## Bayesian Reporting Workflow

The Bayesian approach in ggstatsplot uses the `BayesFactor` package internally. The goal is to quantify *how much* evidence exists for or against an effect, rather than simply crossing a significance threshold.

### Setting priors

```{r}
#| eval: false
library(ggstatsplot)

# Default Cauchy prior (r = 0.707) — appropriate for most comparisons
ggbetweenstats(
  data         = overall_up,
  x            = intervention,
  y            = sbp_0m,
  type         = "bayes",
  bf.prior     = 0.707,        # scale of Cauchy prior on effect size
  title        = "SBP by Intervention (Bayesian, default prior)"
)

# Narrow prior — use if you expect a small effect
ggbetweenstats(
  data         = overall_up,
  x            = intervention,
  y            = sbp_0m,
  type         = "bayes",
  bf.prior     = 0.5,
  title        = "SBP by Intervention (narrow prior)"
)
```

### Reporting Bayes factors in text

When BF₁₀ = 18.4:

> "A Bayesian independent-samples t-test revealed moderate-to-strong evidence
> for a difference in baseline SBP between intervention groups,
> BF₁₀ = 18.4, δ = 0.41, 95% credible interval [0.18, 0.65]."

When BF₁₀ = 0.21:

> "A Bayesian independent-samples t-test provided moderate evidence in favour
> of the null hypothesis (no difference in SBP), BF₁₀ = 0.21."

::: {.callout-tip}
## Key distinction: BF₁₀ vs BF₀₁
- **BF₁₀**: evidence for H₁ over H₀ (use when testing for an effect)
- **BF₀₁ = 1/BF₁₀**: evidence for H₀ over H₁ (use when supporting the null)

ggstatsplot reports BF₁₀ by default. To interpret support for H₀, invert it.
:::

---

## `grouped_*` Functions — Faceted Comparisons

The `grouped_*` variants run the same comparison separately within each level of a grouping variable and combine the results into a faceted plot.

```{r}
#| eval: false
# Compare SBP across tobacco types, separately by sex
grouped_ggbetweenstats(
  data             = overall_up,
  x                = tobacco_type,
  y                = sbp_0m,
  grouping.var     = sex,           # one panel per sex
  title            = "SBP by Tobacco Type, stratified by Sex",
  plotgrid.args    = list(ncol = 2)
)
```

```{r}
#| eval: false
# Correlation stratified by intervention group
grouped_ggscatterstats(
  data         = overall_up,
  x            = current_age,
  y            = sbp_0m,
  grouping.var = intervention,
  title        = "Age vs SBP by Intervention Group"
)
```

**Available grouped functions:**

- `grouped_ggbetweenstats()`
- `grouped_ggwithinstats()`
- `grouped_ggscatterstats()`
- `grouped_gghistostats()`
- `grouped_ggbarstats()`
- `grouped_ggpiestats()`
- `grouped_ggcorrmat()`
- `grouped_ggdotplotstats()`

---

## `ggcorrmat()` — Correlation Matrix

Visualises pairwise correlations across multiple continuous variables with significance annotations.

```{r}
#| eval: false
ggcorrmat(
  data   = overall_up |> select(current_age, sbp_0m, sbp_3m, sbp_6m,
                                 consumption_per_day_0m),
  title  = "Correlation Matrix — Baseline Variables",
  type   = "parametric"     # "parametric", "nonparametric", "robust", "bayes"
)
```

**Choosing the correlation type:**

| `type` | Method | When to use |
|---|---|---|
| `"parametric"` | Pearson's r | Continuous, approximately normal |
| `"nonparametric"` | Spearman's ρ | Ordinal or skewed distributions |
| `"robust"` | Percentage bend r | Presence of outliers |
| `"bayes"` | Bayesian r | Quantify evidence, not just significance |

---

## Combining with patchwork

All ggstatsplot outputs are ggplot2 objects and compose naturally with `patchwork`.

```{r}
#| eval: false
library(patchwork)

p1 <- ggbetweenstats(
  data  = overall_up,
  x     = intervention,
  y     = sbp_0m,
  title = "Baseline SBP"
)

p2 <- ggbetweenstats(
  data  = overall_up,
  x     = intervention,
  y     = sbp_3m,
  title = "3-Month SBP"
)

p3 <- ggscatterstats(
  data  = overall_up,
  x     = sbp_0m,
  y     = sbp_3m,
  title = "Baseline vs 3-Month SBP"
)

# Compose: two plots on top, one full-width below
(p1 | p2) / p3 +
  plot_annotation(
    title   = "Blood Pressure Outcomes by Intervention",
    caption = "Source: NCTOH Study Dataset"
  )
```

---

## Publication-ready output

```{r}
#| eval: false
library(ggplot2)

pub_plot <- ggbetweenstats(
  data             = overall_up,
  x                = intervention,
  y                = sbp_0m,
  type             = "parametric",
  pairwise.display = "significant",    # show only significant comparisons
  p.adjust.method  = "bonferroni",
  title            = "Baseline Systolic Blood Pressure by Intervention Group",
  xlab             = "Intervention Group",
  ylab             = "Systolic BP (mmHg)",
  package          = "RColorBrewer",
  palette          = "Set1"
) +
  theme_classic(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", hjust = 0.5),
    axis.line        = element_line(linewidth = 0.5),
    plot.caption     = element_text(hjust = 0, colour = "grey50")
  ) +
  labs(caption = "Bonferroni-adjusted pairwise comparisons shown for p < 0.05")

# Save at print quality
ggsave("figures/sbp_by_intervention.png",
       pub_plot, width = 8, height = 6, dpi = 300)
```

---

## Case Study: Healthcare Data

A complete example using the NCTOH study dataset — comparing the change in SBP across intervention groups, accounting for tobacco type.

```{r}
#| eval: false
library(ggstatsplot)
library(dplyr)
library(patchwork)

# Derive SBP change variable
study_data <- overall_up |>
  mutate(sbp_change = sbp_3m - sbp_0m)

# 1. Overall change by intervention
p_overall <- ggbetweenstats(
  data  = study_data,
  x     = intervention,
  y     = sbp_change,
  type  = "parametric",
  title = "Change in SBP at 3 Months"
)

# 2. Stratified by tobacco type
p_strat <- grouped_ggbetweenstats(
  data         = study_data,
  x            = intervention,
  y            = sbp_change,
  grouping.var = tobacco_type,
  title        = "SBP Change Stratified by Tobacco Type",
  plotgrid.args = list(ncol = 3)
)

# View each
p_overall
p_strat
```

---

## Quick reference

```{r}
#| eval: false
# Bayesian comparison
ggbetweenstats(data = df, x = group, y = outcome, type = "bayes",
               bf.prior = 0.707)

# Faceted comparisons
grouped_ggbetweenstats(data = df, x = group, y = outcome,
                       grouping.var = stratifier)

# Correlation matrix
ggcorrmat(data = df |> select(var1, var2, var3, var4))

# Compose with patchwork
(p1 | p2) / p3 + plot_annotation(title = "Figure 1")
```
