---
title: "Data Analysis"
subtitle: "Normality Testing, Descriptive Statistics, Parametric and Non-Parametric Tests"
author: "Workshop"
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    toc: false
    chalkboard: false
    incremental: false
    smaller: false
    scrollable: false
execute:
  echo: true
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(readxl)
library(janitor)
library(nortest)
library(car)
library(rstatix)
library(exact2x2)

p <- read_excel("../data/prac_modded.xlsx") |>
  clean_names()
```

---

# Why Data Analysis Matters {background-color="#2C3E50"}

## Importance of Data Analysis

:::: {.columns}

::: {.column width="50%"}
**Making Sense of Data**

- **Driving Informed Decisions** — strategic choices based on evidence, not intuition
- **Identifying Patterns and Insights** — uncovering hidden relationships in data
- **Optimizing Processes** — streamline operations using performance metrics
- **Supporting Innovation** — fuel new ideas and solutions
:::

::: {.column width="50%"}
**Research Perspective**

- **Evidence-Based Research** — validate hypotheses and draw reliable conclusions
- **Mitigating Risk** — identify potential errors before they reach conclusions
- **Improving Reproducibility** — ensures findings can be replicated
- **Ethical Responsibility** — especially critical in clinical and health research
:::

::::

---

# Choosing the Right Statistical Test {background-color="#2C3E50"}

## Why Test Selection Matters

Choosing the wrong test leads to **Type I** (false positive) or **Type II** (false negative) errors.

| Reason | Detail |
|---|---|
| Accuracy & Validity | Test must match data nature and research question |
| Avoiding Errors | Wrong test → misleading conclusions |
| Reproducibility | Correct test ensures others can replicate findings |
| Meeting Objectives | Test must align with the hypothesis |
| Ethical Responsibility | In clinical research, conclusions affect real policy |

---

## Decision Framework — Overview

> The choice of statistical test depends on: **variable type → number of groups → normality → independence of observations**

:::{.callout-note}
**Flowchart placeholder**

*Insert here: Overall test selection flowchart — starting from variable type (Categorical vs Continuous), then branching by number of groups, then normality, then paired vs unpaired.*
:::

---

# Variables {background-color="#2C3E50"}

## Types of Variables

:::{.callout-note}
**Flowchart placeholder**

*Insert here: Variable classification diagram — Qualitative (Nominal, Ordinal) and Quantitative (Discrete, Continuous). Can also show examples relevant to your dataset such as sex, wealth index, age, SBP, FBS.*
:::

:::: {.columns}

::: {.column width="50%"}
**Categorical / Qualitative**

- **Nominal** — no order (e.g., sex, past history of DM)
- **Ordinal** — ordered categories (e.g., wealth index: poorest → richest)
- **Binary / Dichotomous** — two categories (e.g., screened: Yes/No)
:::

::: {.column width="50%"}
**Numerical / Quantitative**

- **Discrete** — whole numbers (e.g., number of visits)
- **Continuous** — any value in a range (e.g., age, SBP, height, FBS)
:::

::::

---

# Descriptive Statistics {background-color="#2C3E50"}

## Categorical Variables — Frequency Tables

Use `table()` for base R frequency counts or `group_by()` + `reframe()` for tidy output.

```{r}
# Base R approach
freq_table <- table(p$past_history_of_diabetes)
prop_table <- prop.table(freq_table)

combined_table <- data.frame(
  Frequency = freq_table,
  Percentage = round(prop_table * 100, 2)
)
combined_table
```

---

## Categorical Variables — Tidyverse Approach

```{r}
# Frequency and percentage using dplyr — unordered factor
summary_table <- p %>%
  mutate(wealth_index = factor(wealth_index_of_participants)) |>
  group_by(wealth_index) %>%
  reframe(
    Frequency = n(),
    Percentage = paste(round(n() * 100 / nrow(p), 2), "%")
  )
summary_table
```

---

## Categorical Variables — Ordered Factor

```{r}
# Define a meaningful order for wealth index
summary_table_1 <- p %>%
  mutate(wealth_index = factor(
    wealth_index_of_participants,
    levels = c("poorest", "poorer", "middle", "richer", "richest")
  )) |>
  group_by(wealth_index) %>%
  reframe(
    Frequency = n(),
    Percentage = paste(round(n() * 100 / nrow(p), 2), "%")
  )
summary_table_1
```

---

## Continuous Variables — Summary Statistics

```{r}
# Base R
summary(p$age_of_participants)
sd(p$age_of_participants)
```

---

## Continuous Variables — Full Tidyverse Summary

```{r}
p |>
  summarise(
    mean_age    = mean(age_of_participants),
    sd_age      = sd(age_of_participants),
    median_age  = median(age_of_participants),
    Q1          = quantile(age_of_participants, 0.25),
    Q3          = quantile(age_of_participants, 0.75),
    IQR         = IQR(age_of_participants),
    var_age     = var(age_of_participants),
    min_age     = min(age_of_participants),
    max_age     = max(age_of_participants)
  )
```

---

# Normality Testing {background-color="#2C3E50"}

## Why Check Normality?

Many parametric tests (t-test, ANOVA, Pearson's r) **assume normally distributed data**.

Failing to check normality can lead to:

- Invalid p-values
- Wrong test selection
- Misleading conclusions

**Two approaches:** Visual methods + Statistical tests

---

## Visual Methods — Histogram

```{r}
hist(
  p$age_of_participants,
  main  = "Histogram with Density Line",
  xlab  = "Age",
  ylab  = "Density",
  col   = "grey",
  border = "black",
  freq  = FALSE
)
lines(density(p$age_of_participants), col = "red", lwd = 2)
```

*A roughly bell-shaped curve suggests normality. Skewed or multimodal histograms indicate non-normality.*

---

## Visual Methods — QQ Plot

```{r}
qqnorm(p$sbp_in_mm_hg, main = "Q-Q Plot: SBP")
qqline(p$sbp_in_mm_hg, col = "red")
```

*Points should follow the red diagonal line closely if the data are normally distributed. Deviations at the tails indicate non-normality.*

---

## Statistical Tests — Shapiro-Wilk

Best for **small samples (n ≤ 50)**. Randomly sample if n > 50.

```{r}
p1 <- p[sample(nrow(p), 50), ]
s_test <- shapiro.test(p1$age_of_participants)
s_test
```

> **Interpretation:** p > 0.05 → data does not significantly deviate from normality (cannot reject H₀ of normality).

---

## Statistical Tests — Kolmogorov-Smirnov

Suitable for **larger samples**. Compares distribution to a theoretical normal.

```{r}
ks_test <- ks.test(p$sbp_in_mm_hg, "pnorm",
                   mean = mean(p$sbp_in_mm_hg),
                   sd   = sd(p$sbp_in_mm_hg))
ks_test
```

> **Note:** Use `options(scipen = 999)` to display p-values in decimal format instead of scientific notation.

---

## Statistical Tests — Anderson-Darling

More sensitive to tails than KS test. Requires `nortest` package.

```{r}
ad_test <- ad.test(p$height)
ad_test
```

| Test | Best Use | Package |
|---|---|---|
| Shapiro-Wilk | n ≤ 50 | base R |
| Kolmogorov-Smirnov | n > 50, continuous | base R |
| Anderson-Darling | Sensitive to tail deviations | `nortest` |

---

# Statistical Tests — Categorical Variables {background-color="#2C3E50"}

## Decision Framework — Categorical

:::{.callout-note}
**Flowchart placeholder**

*Insert here: Categorical variable test selection flowchart — branching from number of groups (1, 2, >2) and paired vs unpaired, leading to: Chi-square goodness of fit → Chi-square test of independence / Fisher's exact → McNemar / Exact McNemar → Chi-square m×n / Monte Carlo.*
:::

---

## Chi-Square Goodness-of-Fit — 1 Group

Tests whether the observed distribution matches an **expected distribution**.

```{r}
observed <- table(p$ever_screened_for_bp)
total    <- sum(observed)
expected <- c(1/2, 1/2) * total

chisq_test <- chisq.test(x = observed, p = expected / sum(expected))
chisq_test
```

> **Assumption:** Expected frequency ≥ 5 in each cell.

---

## Chi-Square Test — 2 Binary Groups (Expected ≥ 5)

Tests association between two categorical variables using a **contingency table**.

```{r}
table_gender_dm <- table(p$sex_of_participants, p$past_history_of_diabetes)
table_gender_dm

chisq_m5 <- chisq.test(table_gender_dm, correct = FALSE)
chisq_m5

# Check expected frequencies
chisq_m5$expected
```

> Use `correct = FALSE` to apply the standard Pearson Chi-square (no Yates correction).

---

## Chi-Square — Yates Continuity Correction

Yates correction subtracts 0.5 from the absolute difference between observed and expected frequency before squaring.

```{r}
# With Yates correction (correct = TRUE is default for 2x2 tables)
chisq.test(table_gender_dm, correct = TRUE)
```

> Apply when sample sizes are small or expected frequencies approach 5.

---

## When Expected Frequency < 5

Two options: **Monte Carlo simulation** or **Fisher's Exact Test**.

```{r}
table_htn_dm <- table(p$ever_screened_for_bp, p$past_history_of_diabetes)

# Monte Carlo simulation (B = number of replicates, min p = 1/(B+1))
chisq_mc <- chisq.test(table_htn_dm, simulate.p.value = TRUE, B = 2000)
chisq_mc

# Fisher's Exact Test
fisher_test <- fisher.test(table_htn_dm)
fisher_test
```

---

## Chi-Square — m × n Contingency Table

Use when both variables have **more than two categories**.

```{r}
table_mar_wealth <- table(
  p$marital_status_of_participants,
  p$wealth_index_of_participants
)
table_mar_wealth

chisq_mn <- chisq.test(table_mar_wealth)
chisq_mn

# Monte Carlo if expected frequencies < 5
chisq_mn_mc <- chisq.test(table_mar_wealth, simulate.p.value = TRUE, B = 2000)
chisq_mn_mc
```

---

## McNemar Test — Paired Binary Responses

Use when the **same participants** are measured twice on a binary outcome (e.g., before and after treatment).

```{r}
table_bfr_aftr <- table(p$pain_foot_bfr, p$pain_foot_aftr)
table_bfr_aftr

mcnemar_test <- mcnemar.test(table_bfr_aftr)
mcnemar_test
```

> **Assumption:** Discordant cells (b + c) should be ≥ 25. If < 25, use exact McNemar.

---

## Exact McNemar Test

```{r}
exact_mcnemar <- exact2x2::mcnemar.exact(table_bfr_aftr)
exact_mcnemar
```

| Situation | Test |
|---|---|
| Paired binary, discordant ≥ 25 | McNemar |
| Paired binary, discordant < 5 | Exact McNemar (`exact2x2`) |

---

# Statistical Tests — Continuous Variables {background-color="#2C3E50"}

## Decision Framework — Continuous

:::{.callout-note}
**Flowchart placeholder**

*Insert here: Continuous variable test selection flowchart — branching from number of groups (1, 2, >2), then normality (normal vs non-normal), then paired vs unpaired, leading to: Z-test / One-sample t → Independent t / Paired t / Mann-Whitney U / Wilcoxon Matched Pairs → One-way ANOVA / Repeated Measures ANOVA / Kruskal-Wallis / Friedman.*
:::

---

## One Group — Z Test

Use when population SD is **known** and n > 30.

```{r}
pr  <- p |> filter(wealth_index_of_participants == "richer")
mu  <- mean(p$age_of_participants)
sigma <- sd(p$age_of_participants)

s_mean <- mean(pr$age_of_participants)
n      <- length(pr$age_of_participants)
z      <- (s_mean - mu) / (sigma / sqrt(n))

p_two_tailed     <- 2 * (1 - pnorm(abs(z)))
p_one_tailed_gt  <- 1 - pnorm(z)
p_one_tailed_lt  <- pnorm(z)

cat("Z =", round(z, 3),
    "\nTwo-tailed p =", round(p_two_tailed, 4),
    "\nOne-tailed (greater) p =", round(p_one_tailed_gt, 4))
```

---

## One Group — One-Sample t Test

Use when population SD is **unknown** or n < 30.

```{r}
pp <- p |> filter(wealth_index_of_participants == "poorest")

t_test <- t.test(pp$age_of_participants, mu = mu)
t_test
```

| Test | Population SD | n |
|---|---|---|
| One-sample Z | Known | > 30 |
| One-sample t | Unknown | Any; especially < 30 |

---

## Two Groups — Testing Equal Variance

Before an independent t-test, verify the **equal variance assumption**.

```{r}
# Levene's test (robust to non-normality)
car::leveneTest(weight ~ sex_of_participants, data = p)

# Bartlett's test (assumes normality)
bartlett.test(weight ~ sex_of_participants, data = p)
```

> **Interpretation:** p < 0.05 → variances are NOT equal → use `var.equal = FALSE` in t.test.

---

## Two Groups — Independent t Test (Unpaired, Normal)

```{r}
ind_t_test <- t.test(weight ~ sex_of_participants,
                     var.equal = TRUE,
                     data = p)
ind_t_test
```

:::{.callout-note}
**Flowchart placeholder**

*Insert here: Equal variance flowchart — Levene/Bartlett test → p < 0.05 (unequal variances, Welch t-test) vs p ≥ 0.05 (equal variances, standard independent t-test).*
:::

---

## Two Groups — Paired t Test (Normal)

Use when the **same participants** are measured at two time points.

```{r}
paired_t_test <- t.test(p$fbs_presentation, p$fbs_6_months, paired = TRUE)
paired_t_test
```

> **Assumption:** Differences between paired observations must be normally distributed.

---

## Two Groups — Mann-Whitney U (Non-Normal, Unpaired)

Non-parametric alternative to the independent t-test. Based on **ranks**, not raw values.

```{r}
mann_whit <- wilcox.test(sbp_presentation ~ sex_of_participants,
                         data = p,
                         correct = FALSE)
mann_whit
```

> Use `correct = TRUE` (continuity correction) when: sample size < 100, or data is ordinal/discrete.

---

## Two Groups — Wilcoxon Matched Pairs (Non-Normal, Paired)

Non-parametric alternative to the paired t-test.

```{r}
wilcox_paired <- wilcox.test(p$sbp_presentation, p$sbp_6_months, paired = TRUE)
wilcox_paired
```

> **Assumption:** Differences between paired values should be **symmetrically distributed** (not necessarily normal).

---

## Summary — Two Group Tests

| Groups | Normal | Test |
|---|---|---|
| Unpaired | Yes | Independent t-test |
| Unpaired | No | Mann-Whitney U |
| Paired | Yes | Paired t-test |
| Paired | No | Wilcoxon Matched Pairs |

---

# More Than Two Groups {background-color="#2C3E50"}

## One-Way ANOVA (Normal, Unpaired)

Tests for differences in means across **three or more independent groups**.

```{r}
aov_un <- aov(height ~ marital_status_of_participants, data = p)
summary(aov_un)
```

> **Assumptions:** Groups are independent, data within each group is normally distributed, equal variances across groups.

---

## One-Way ANOVA — Post Hoc (Tukey HSD)

A significant ANOVA tells you that at least one group differs — but not **which pairs**.

```{r}
tukey_res <- TukeyHSD(aov_un)
print(tukey_res)
```

> Tukey's HSD controls the family-wise error rate when making all pairwise comparisons.

---

## Repeated Measures ANOVA (Normal, Paired)

Use when the **same participants** are measured at three or more time points.

```{r}
p_long_n <- p |>
  select(s_no, fbs_presentation, fbs_3_months, fbs_6_months) |>
  pivot_longer(
    cols = c(fbs_presentation, fbs_3_months, fbs_6_months),
    names_to = "time",
    values_to = "value"
  )

res_aov <- anova_test(data = p_long_n, dv = value, wid = s_no, within = time)
get_anova_table(res_aov)
```

> **Assumption:** Sphericity — variances of differences between all pairs of time points are equal.

---

## Repeated Measures ANOVA — Post Hoc

```{r}
pairwise_p <- pairwise.t.test(
  p_long_n$value,
  p_long_n$time,
  p.adjust.method = "bonferroni"
)
print(pairwise_p)
```

> **Bonferroni correction** divides the significance level (α) by the number of comparisons to control Type I error.

---

## Kruskal-Wallis Test (Non-Normal, Unpaired)

Non-parametric alternative to one-way ANOVA. Based on **ranks**.

```{r}
kruskal_unp <- kruskal.test(sbp_presentation ~ marital_status_of_participants, data = p)
kruskal_unp
```

---

## Kruskal-Wallis — Post Hoc

```{r}
pairwise_wilcox <- pairwise.wilcox.test(
  p$sbp_presentation,
  p$marital_status_of_participants,
  p.adjust.method = "bonferroni"
)
print(pairwise_wilcox)
```

> **Minimum achievable p-value** with Monte Carlo simulation of B replicates = 1 / (B + 1).

---

## Friedman Test (Non-Normal, Paired / Repeated)

Non-parametric equivalent of repeated measures ANOVA.

```{r}
data_long <- pivot_longer(
  p,
  cols = c(sbp_presentation, sbp_3_months, sbp_6_months),
  names_to  = "time",
  values_to = "sbp"
)

friedman_res <- friedman.test(sbp ~ time | s_no, data = data_long)
print(friedman_res)
```

> **Assumption:** Dependent samples — same participants measured at each time point.

---

## Friedman — Post Hoc

```{r}
pairwise_np <- pairwise.wilcox.test(
  data_long$sbp,
  data_long$time,
  p.adjust.method = "bonferroni",
  paired = TRUE
)
print(pairwise_np)
```

---

## Summary — More Than Two Groups

| Groups | Normal | Design | Test | Post Hoc |
|---|---|---|---|---|
| > 2 | Yes | Unpaired | One-way ANOVA | Tukey HSD |
| > 2 | Yes | Paired | Repeated ANOVA | Bonferroni t-test |
| > 2 | No | Unpaired | Kruskal-Wallis | Pairwise Wilcoxon |
| > 2 | No | Paired | Friedman | Pairwise Wilcoxon |

---

# Correlation {background-color="#2C3E50"}

## Pearson vs Spearman Correlation

Both measure **strength and direction** of a relationship between two continuous variables.

| Feature | Pearson | Spearman |
|---|---|---|
| Data requirement | Normally distributed | Any distribution |
| Relationship type | Linear | Monotonic |
| Homoscedasticity | Required | Not required |
| Based on | Raw values | Ranks |

> **Homoscedasticity** — the variance of residuals should remain constant across all levels of the independent variable.

---

## Pearson's Correlation

```{r}
pearson <- cor.test(p$fbs_presentation, p$weight, method = "pearson")
pearson
```

> Assumes: linearity, bivariate normality, homoscedasticity, independence of observations.

---

## Spearman's Correlation

```{r}
spearman <- cor.test(
  p$sbp_in_mm_hg,
  p$age_of_participants,
  method = "spearman",
  exact = FALSE
)
spearman
```

> Use `exact = FALSE` when there are ties in the data (avoids exact p-value computation error).

---

## Interpreting Correlation Coefficients

| r value | Interpretation |
|---|---|
| 0.00 – 0.19 | Negligible |
| 0.20 – 0.39 | Weak |
| 0.40 – 0.59 | Moderate |
| 0.60 – 0.79 | Strong |
| 0.80 – 1.00 | Very strong |

> Always report both **r** (strength and direction) and **p** (statistical significance) together.

---

## Overall Test Selection — Full Decision Map

:::{.callout-note}
**Flowchart placeholder**

*Insert here: Master decision flowchart — starting from variable type (Categorical vs Continuous), then number of groups (1, 2, >2), then normality check, then paired vs unpaired, with all tests mapped to the correct branch. This is the key summary diagram for the whole session.*
:::

---

# Cleaning Up the Environment

```{r}
rm(list = setdiff(ls(), "p"))
```

> Use `options(scipen = 999)` to display p-values as decimals instead of scientific notation.
> Use `options(scipen = 0)` to revert to scientific notation.

---

# Coming Next {background-color="#2C3E50"}

## Further Topics

- **Data Visualization** — principles of effective charts
- **ggplot2** — grammar of graphics in R
- **ggstatsplot** — publication-ready plots with embedded statistics
