---
title: "Data Analysis"
format:
  html:
    code-copy: true
    theme: cosmo
    toc: true
    lightbox: auto
---

::: {.content-visible when-profile="moderate"}

# Data Analysis for Medical Datasets

This section builds on general R skills and shows common descriptive and inferential analyses.

```{r}
library(tidyverse)
library(readxl)
library(janitor)
```

## Load and clean data

```{r}
p <- read_excel("../data/prac_modded.xlsx") |> 
  clean_names()

head(p)
```

## Descriptive statistics

```{r}
summary(p$age_of_participants)
sd(p$age_of_participants)
```

### Frequency tables

```{r}
freq_table <- table(p$past_history_of_diabetes)
prop_table <- prop.table(freq_table)
combined_table <- data.frame(
  Frequency = freq_table,
  Percentage = prop_table * 100
)
combined_table
```

### Grouped summaries

```{r}
summary_table <- p %>%
  mutate(wealth_index = factor(wealth_index_of_participants)) %>% 
  group_by(wealth_index) %>%
  reframe(
    Frequency = n(),
    Percentage = paste0(round(n() * 100 / nrow(p), 2), "%")
  )
summary_table
```

## Hypothesis testing

### Chi-square goodness-of-fit

```{r}
observed <- table(p$ever_screened_for_bp)
expected <- c(1/2, 1/2) * sum(observed)
chisq_test <- chisq.test(x = observed, p = expected / sum(observed))
chisq_test
```

### Chi-square test for association

```{r}
table_gender_dm <- table(p$sex_of_participants, p$past_history_of_diabetes)
chisq_m_5_b <- chisq.test(table_gender_dm, correct = FALSE)
chisq_m_5_b
```

### McNemar test for paired binary responses

```{r}
table_bfr_aftr <- table(p$pain_foot_bfr, p$pain_foot_aftr)
mcnemar_test <- mcnemar.test(table_bfr_aftr)
mcnemar_test
```

### Normality diagnostics

```{r}
hist(p$age_of_participants, main = "Age distribution", xlab = "Age", ylab = "Density", freq = FALSE, col = "grey")
lines(density(p$age_of_participants), col = "red", lwd = 2)
qqnorm(p$sbp_in_mm_hg)
qqline(p$sbp_in_mm_hg, col = "red")
```

These examples show the R workflow for medical data: import, clean, summarize, test, and visualize.

:::

::: {.content-visible when-profile="advanced"}

## Automating Analysis with Loops

Real datasets often need the same test or summary repeated across many variables. Loops (and their `apply`/`purrr` equivalents) let you do that without copy-pasting code.

### `for` loops

```{r}
vars <- c("age_of_participants", "sbp_in_mm_hg", "dbp_in_mm_hg")

for (v in vars) {
  cat("Summary of", v, ":\n")
  print(summary(p[[v]]))
  cat("\n")
}
```

### `while` loops

```{r}
# Halve a sample size threshold until it drops below 10
n <- nrow(p)
threshold <- n

while (threshold > 10) {
  threshold <- threshold %/% 2
}

cat("Final threshold:", threshold, "\n")
```

### `repeat` loops

```{r}
# repeat runs until an explicit break condition is met
count <- 0
repeat {
  count <- count + 1
  if (count >= 5) break
}
count
```

### Vectorized alternatives: `sapply()` and `purrr::map()`

Loops are explicit and easy to read, but for simple per-column operations the `apply` family or `purrr::map()` are usually shorter and avoid manual bookkeeping:

```{r}
library(purrr)

# Same summaries as the for-loop above, without the explicit loop
sapply(p[vars], mean, na.rm = TRUE)

map_dbl(p[vars], ~ sd(.x, na.rm = TRUE))
```

Use a `for` loop when each iteration has side effects (printing, saving files, building up a report) or when the logic is easier to follow step-by-step; reach for `sapply()`/`map()` when you just need one result per element.

:::
