---
title: "ggstatsplot — Introduction"
format:
  html:
    code-copy: true
    toc: true
    lightbox: auto
---

## What is ggstatsplot?

`ggstatsplot` is an R package that combines **statistical tests** and **publication-ready visualisations** in a single function call. Each plot automatically selects, runs, and annotates the appropriate statistical test based on your data, displaying results directly on the figure.

This removes a common workflow bottleneck: running a test in one place, building a plot in another, and manually annotating results.

```{r}
#| eval: false
install.packages("ggstatsplot")
library(ggstatsplot)
```

**Why use it?**

- One function produces both the visual and the statistical result
- Automatically annotates effect sizes and confidence intervals
- Sensible defaults — works out of the box for most common comparisons
- Output is a ggplot2 object, so you can add layers or themes as usual

---

## Core functions covered in this module

| Function | Use case |
|---|---|
| `ggbetweenstats()` | Compare groups (independent samples) |
| `ggscatterstats()` | Correlation between two continuous variables |
| `gghistostats()` | Distribution of a single continuous variable |

---

## `ggbetweenstats()` — Group Comparisons

Use this when you want to compare a continuous variable across two or more independent groups.

```{r}
#| eval: false
library(ggstatsplot)

# Compare systolic BP across fuel type groups
ggbetweenstats(
  data  = overall_up,
  x     = fuel_type,         # grouping variable (categorical)
  y     = sbp_0m,            # outcome variable (continuous)
  title = "Systolic BP by Fuel Type"
)
```

**What the output shows:**

- A violin + boxplot combination with individual data points
- The statistical test used (Welch's ANOVA or Kruskal-Wallis, chosen automatically)
- The p-value, effect size (η² or ε²), and confidence interval
- Pairwise comparisons (if more than 2 groups)

**Choosing parametric vs non-parametric:**

```{r}
#| eval: false
# Force non-parametric (Kruskal-Wallis)
ggbetweenstats(
  data  = overall_up,
  x     = fuel_type,
  y     = sbp_0m,
  type  = "nonparametric"   # "parametric", "nonparametric", "robust", "bayes"
)
```

---

## `ggscatterstats()` — Correlation

Use this when you want to examine the relationship between two continuous variables.

```{r}
#| eval: false
ggscatterstats(
  data  = overall_up,
  x     = current_age,
  y     = sbp_0m,
  title = "Age vs Systolic BP at Baseline"
)
```

**What the output shows:**

- A scatter plot with a linear trend line and confidence band
- Pearson (or Spearman) correlation coefficient, r
- p-value and 95% confidence interval for r
- Marginal histograms on both axes

**Using Spearman instead of Pearson:**

```{r}
#| eval: false
ggscatterstats(
  data = overall_up,
  x    = current_age,
  y    = sbp_0m,
  type = "nonparametric"   # uses Spearman's rho
)
```

---

## `gghistostats()` — Distribution

Use this when you want to visualise the distribution of a single continuous variable with a one-sample test.

```{r}
#| eval: false
gghistostats(
  data     = overall_up,
  x        = current_age,
  title    = "Distribution of Age at Enrolment",
  test.value = 40          # hypothesised population mean
)
```

**What the output shows:**

- Histogram with overlaid density curve
- One-sample t-test (or Wilcoxon signed-rank) against the specified `test.value`
- Mean, median, and their confidence intervals

---

## Reading the statistical annotation

Every ggstatsplot figure includes a subtitle with the test results in a standardised format:

```
Welch's ANOVA, F(2, 147.3) = 5.23, p = 0.006, ω² = 0.06, 95% CI [0.01, 0.14]
```

| Component | Meaning |
|---|---|
| Test name | Which test was used (selected automatically) |
| Test statistic | F, t, χ², H etc. |
| p-value | Statistical significance |
| Effect size | Practical significance (ω², d, r, etc.) |
| 95% CI | Uncertainty around the effect size |

::: {.callout-tip}
## Interpreting effect sizes

- **η² / ω²**: Small ≈ 0.01, Medium ≈ 0.06, Large ≈ 0.14 (for ANOVA)
- **Cohen's d**: Small ≈ 0.2, Medium ≈ 0.5, Large ≈ 0.8 (for t-tests)
- **r**: Small ≈ 0.1, Medium ≈ 0.3, Large ≈ 0.5 (for correlation)
:::

---

## Quick reference

```{r}
#| eval: false
# Two-group comparison
ggbetweenstats(data = df, x = group, y = outcome)

# Correlation
ggscatterstats(data = df, x = var1, y = var2)

# Distribution
gghistostats(data = df, x = continuous_var, test.value = 0)
```

---

## Further reading

- [ggstatsplot documentation](https://indrajeetpatil.github.io/ggstatsplot/)
- Patil, I. (2021). Visualizations with statistical details. *JOSS*, 6(61), 3167.
