---
title: "Quarto — Dynamic Documents"
format:
  html:
    code-copy: true
    toc: true
    lightbox: auto
---

::: {.callout-note}
This module builds on [Quarto Basics](reproducible_writing.qmd). It assumes you
can create and render a `.qmd` file and write basic code chunks.
:::

## What you will learn

- Parameterised reports — run the same document for different subgroups
- Inline R code — embed values directly in prose
- Cross-references for figures and tables
- Multiple output formats from one source file
- Integrating `gtsummary` tables in Quarto

---

## Parameterised Reports

Parameters let you run the same document with different inputs — for example,
generating one report per hospital, intervention group, or region.

### 1. Declare parameters in the YAML header

```yaml
---
title: "Site Report"
params:
  site_name: "Site A"
  intervention: "Control"
---
```

### 2. Use parameters in code

```{r}
#| eval: false
# params$site_name and params$intervention are available as R objects
site_data <- overall_up |>
  filter(site == params$site_name,
         intervention == params$intervention)

nrow(site_data)
```

### 3. Render with different parameter values

```{r}
#| eval: false
# Render from R for a specific site
quarto::quarto_render(
  input  = "site_report.qmd",
  output_file = "site_a_report.html",
  execute_params = list(
    site_name    = "Site A",
    intervention = "Intervention"
  )
)
```

### 4. Batch render — one file per site

```{r}
#| eval: false
library(purrr)

sites <- c("Site A", "Site B", "Site C")

walk(sites, \(site) {
  quarto::quarto_render(
    input          = "site_report.qmd",
    output_file    = paste0(tolower(gsub(" ", "_", site)), "_report.html"),
    execute_params = list(site_name = site)
  )
})
```

---

## Inline R Code

Embedding computed values directly in text avoids copy-paste errors and keeps
your document reproducible.

**Syntax:**

`` `r nrow(gtsummary::trial)` ``

```{r}
#| eval: false
# In a code chunk (runs silently):
n_total  <- nrow(overall_up)
mean_age <- round(mean(overall_up$current_age, na.rm = TRUE), 1)
```

**In Quarto prose:**

```markdown
The study enrolled `r nrow(gtsummary::trial)` participants with a mean age of
`r mean(gtsummary::trial$age)` years (SD = `r round(sd(gtsummary::trial$age, na.rm = TRUE), 1)`).
```

Rendered output:

> The study enrolled 1 248 participants with a mean age of 38.4 years (SD = 11.2).

::: {.callout-tip}
Compute all summary values in one early code chunk marked `#| include: false`,
then reference them via inline code throughout the document.
:::

---

## Cross-References

Cross-references auto-number figures and tables and create clickable links.

### Figures

Add a `label` that starts with `fig-` and a `fig-cap`:

````markdown
```{{r}}
#| label: fig-sbp-hist
#| fig-cap: "Distribution of baseline systolic blood pressure."
ggplot(overall_up, aes(x = sbp_0m)) +
  geom_histogram(bins = 30) +
  labs(x = "Systolic BP (mmHg)", y = "Count")
```
````

Reference in text: `@fig-sbp-hist` → renders as "Figure 1"

### Tables

Add a `label` starting with `tbl-` and a `tbl-cap`:

````markdown
```{{r}}
#| label: tbl-demographics
#| tbl-cap: "Baseline characteristics of study participants."
library(gtsummary)
overall_up |>
  select(current_age, sex, tobacco_type, fuel_type) |>
  tbl_summary()
```
````

Reference in text: `@tbl-demographics` → renders as "Table 1"

---

## Multiple Output Formats

One `.qmd` file can produce HTML, PDF, and Word output.

```yaml
---
title: "Study Report"
format:
  html:
    toc: true
    theme: flatly
  pdf:
    toc: true
    number-sections: true
  docx:
    toc: true
---
```

Render all formats at once:

```bash
quarto render study_report.qmd
```

::: {.callout-warning}
## PDF requirements
PDF rendering requires a LaTeX installation. Install `tinytex` from R:
```r
install.packages("tinytex")
tinytex::install_tinytex()
```
:::

---

## gtsummary Tables in Quarto

`gtsummary` integrates naturally with Quarto. Use `as_kable_extra()` or the
default `print()` method for HTML output; use `as_gt()` for PDF.

```{r}
#| eval: false
#| label: tbl-summary
#| tbl-cap: "Baseline characteristics by intervention group."
library(gtsummary)

overall_up |>
  select(current_age, sex, tobacco_type, fuel_type, sbp_0m, intervention) |>
  tbl_summary(
    by      = intervention,
    missing = "no"
  ) |>
  add_p() |>
  add_overall() |>
  bold_labels()
```

**Adjusting output by format:**

```{r}
#| eval: false
tbl <- overall_up |>
  tbl_summary(by = intervention)

# HTML output
tbl

# For PDF (via gt)
tbl |> as_gt()

# For Word
tbl |> as_flex_table()
```

---

## Code Chunk Options Reference

| Option | Effect |
|---|---|
| `echo: false` | Hide code, show output |
| `include: false` | Run code silently (no output shown) |
| `eval: false` | Show code but don't run it |
| `warning: false` | Suppress warnings |
| `message: false` | Suppress package messages |
| `fig-width: 8` | Figure width in inches |
| `fig-height: 5` | Figure height in inches |
| `fig-dpi: 300` | Resolution for saved figures |

Set globally in YAML to apply to all chunks:

```yaml
execute:
  echo: false
  warning: false
  message: false
```
