---
title: "Regression Analysis"
format:
  html:
    code-copy: true
    theme: cosmo
    toc: true
    lightbox: auto
---

Regression models describe and quantify the relationship between an outcome and one or more predictors. This module builds up in tiers — start with a single predictor, then add predictors, then move on to diagnostics and logistic regression.

## Simple Linear Regression

```{r}
library(tidyverse)

data <- data.frame(
  age = rnorm(100, 50, 10),
  blood_pressure = rnorm(100, 120, 15),
  treatment = sample(c("A", "B"), 100, replace = TRUE)
)

model_simple <- lm(blood_pressure ~ age, data = data)
summary(model_simple)
```

```{r}
ggplot(data, aes(x = age, y = blood_pressure)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Blood Pressure vs Age",
    x = "Age", y = "Blood Pressure"
  )
```

`geom_smooth(method = "lm")` draws the same fitted line as `lm()`, with a shaded confidence band around it.

::: {.content-visible when-profile="moderate"}

## Multiple Regression

Adding more predictors lets the model account for other sources of variation at the same time.

```{r}
library(gtsummary)

model <- lm(blood_pressure ~ age + treatment, data = data)
summary(model)

# Using gtsummary for a publication-ready table
tbl_regression(model)
```

Each coefficient is now interpreted "holding the other predictors constant" — the `treatment` effect is adjusted for `age`, and vice versa.

:::

::: {.content-visible when-profile="advanced"}

## Multiple Regression

Adding more predictors lets the model account for other sources of variation at the same time.

```{r}
library(gtsummary)

model <- lm(blood_pressure ~ age + treatment, data = data)
summary(model)

# Using gtsummary for a publication-ready table
tbl_regression(model)
```

Each coefficient is now interpreted "holding the other predictors constant" — the `treatment` effect is adjusted for `age`, and vice versa.

## Regression Diagnostics

A fitted model is only trustworthy once its assumptions have been checked.

```{r}
par(mfrow = c(2, 2))
plot(model)
```

```{r}
# Residuals vs fitted, and a normal Q-Q check
qqnorm(residuals(model))
qqline(residuals(model), col = "red")
```

Watch for: non-random patterns in residuals-vs-fitted (non-linearity), points far off the Q-Q line (non-normal residuals), and high-leverage points in the residuals-vs-leverage panel.

## Logistic Regression

When the outcome is binary rather than continuous, use `glm()` with a binomial family instead of `lm()`.

```{r}
data$outcome <- sample(c("Improved", "Not Improved"), 100, replace = TRUE)
data$outcome_bin <- ifelse(data$outcome == "Improved", 1, 0)

model_log <- glm(outcome_bin ~ age + treatment, data = data, family = binomial)
summary(model_log)

# Odds ratios instead of raw log-odds
tbl_regression(model_log, exponentiate = TRUE)
```

`exponentiate = TRUE` converts the log-odds coefficients into odds ratios, which are usually easier to report and interpret.

:::
