---
title: "Regression Analysis"
format:
  html:
    code-copy: true
    theme: cosmo
    toc: true
    lightbox: auto
---

Regression models describe and quantify the relationship between an outcome and one or more predictors. This module builds up in tiers — start with a single predictor, then add predictors, then move on to diagnostics and logistic regression.

## Simple Linear Regression

```{r}
library(tidyverse)

data <- data.frame(
  age = rnorm(100, 50, 10),
  blood_pressure = rnorm(100, 120, 15),
  treatment = sample(c("A", "B"), 100, replace = TRUE)
)

model_simple <- lm(blood_pressure ~ age, data = data)
summary(model_simple)
```

```{r}
ggplot(data, aes(x = age, y = blood_pressure)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Blood Pressure vs Age",
    x = "Age", y = "Blood Pressure"
  )
```

`geom_smooth(method = "lm")` draws the same fitted line as `lm()`, with a shaded confidence band around it.

::: {.content-visible when-profile="moderate"}

## Multiple Regression

Adding more predictors lets the model account for other sources of variation at the same time.

```{r}
library(gtsummary)

model <- lm(blood_pressure ~ age + treatment, data = data)
summary(model)

# Using gtsummary for a publication-ready table
tbl_regression(model)
```

Each coefficient is now interpreted "holding the other predictors constant" — the `treatment` effect is adjusted for `age`, and vice versa.

:::

::: {.content-visible when-profile="advanced"}

## Multiple Regression

Adding more predictors lets the model account for other sources of variation at the same time.

```{r}
library(gtsummary)

model <- lm(blood_pressure ~ age + treatment, data = data)
summary(model)

# Using gtsummary for a publication-ready table
tbl_regression(model)
```

Each coefficient is now interpreted "holding the other predictors constant" — the `treatment` effect is adjusted for `age`, and vice versa.

## Regression Diagnostics

A fitted model is only trustworthy once its assumptions have been checked.

```{r}
par(mfrow = c(2, 2))
plot(model)
```

```{r}
# Residuals vs fitted, and a normal Q-Q check
qqnorm(residuals(model))
qqline(residuals(model), col = "red")
```

Watch for: non-random patterns in residuals-vs-fitted (non-linearity), points far off the Q-Q line (non-normal residuals), and high-leverage points in the residuals-vs-leverage panel.

## Logistic Regression

When the outcome is binary rather than continuous, use `glm()` with a binomial family instead of `lm()`.

```{r}
data$outcome <- sample(c("Improved", "Not Improved"), 100, replace = TRUE)
data$outcome_bin <- ifelse(data$outcome == "Improved", 1, 0)

model_log <- glm(outcome_bin ~ age + treatment, data = data, family = binomial)
summary(model_log)

# Odds ratios instead of raw log-odds
tbl_regression(model_log, exponentiate = TRUE)
```

`exponentiate = TRUE` converts the log-odds coefficients into odds ratios, which are usually easier to report and interpret.

:::

::: {.content-visible when-profile="regression"}

## Multiple Regression

Adding more predictors lets the model account for other sources of variation at the same time.

```{r}
library(gtsummary)

model <- lm(blood_pressure ~ age + treatment, data = data)
summary(model)

# Using gtsummary for a publication-ready table
tbl_regression(model)
```

Each coefficient is now interpreted "holding the other predictors constant" — the `treatment` effect is adjusted for `age`, and vice versa.

## Regression Diagnostics

A fitted model is only trustworthy once its assumptions have been checked.

```{r}
par(mfrow = c(2, 2))
plot(model)
```

```{r}
# Residuals vs fitted, and a normal Q-Q check
qqnorm(residuals(model))
qqline(residuals(model), col = "red")
```

Watch for: non-random patterns in residuals-vs-fitted (non-linearity), points far off the Q-Q line (non-normal residuals), and high-leverage points in the residuals-vs-leverage panel.

## Logistic Regression

When the outcome is binary rather than continuous, use `glm()` with a binomial family instead of `lm()`.

```{r}
data$outcome <- sample(c("Improved", "Not Improved"), 100, replace = TRUE)
data$outcome_bin <- ifelse(data$outcome == "Improved", 1, 0)

model_log <- glm(outcome_bin ~ age + treatment, data = data, family = binomial)
summary(model_log)

# Odds ratios instead of raw log-odds
tbl_regression(model_log, exponentiate = TRUE)
```

`exponentiate = TRUE` converts the log-odds coefficients into odds ratios, which are usually easier to report and interpret.

## Univariate Regression Screening

Before building a multivariable model, it's common to screen each predictor on its own first. `gtsummary::tbl_uvregression()` fits one regression per predictor and lays the results out in a single comparison table.

```{r}
data$bmi <- rnorm(100, 27, 4)

tbl_uvregression(
  data[, c("blood_pressure", "age", "treatment", "bmi")],
  method = lm,
  y = blood_pressure,
  hide_n = TRUE
)
```

Each row is its own simple regression, fit independently — this is a screening step, not a substitute for the multivariable model above, since it ignores confounding between predictors.

## Cox Proportional Hazards Regression

For time-to-event outcomes (e.g. time to recovery, time to relapse), use `survival::coxph()` instead of `lm()`/`glm()`. It models the *hazard* of the event happening at any given time, adjusted for covariates.

```{r}
library(survival)

data$time  <- rexp(100, rate = 0.05)
data$event <- rbinom(100, 1, 0.7)   # 1 = event observed, 0 = censored

cox_model <- coxph(Surv(time, event) ~ age + treatment, data = data)
summary(cox_model)

# Hazard ratios instead of raw log-hazard coefficients
tbl_regression(cox_model, exponentiate = TRUE)
```

`exponentiate = TRUE` reports hazard ratios: above 1 means a faster time-to-event (higher instantaneous risk) as the predictor increases, below 1 means a slower one.

```{r}
fit <- survfit(Surv(time, event) ~ treatment, data = data)
plot(
  fit, col = c("#3498db", "#e74c3c"), lwd = 2,
  xlab = "Time", ylab = "Survival probability",
  main = "Kaplan-Meier Curves by Treatment"
)
legend("topright", legend = c("Treatment A", "Treatment B"),
       col = c("#3498db", "#e74c3c"), lwd = 2)
```

The Kaplan-Meier curve shows the estimated probability of not yet having experienced the event over time for each treatment group — a purely descriptive view, independent of the Cox model's covariate adjustment above.

## Model Diagnostics with `performance`

The `performance` package gives a quick, standardized diagnostic check for a fitted model, going further than the base-R diagnostic plots used earlier.

```{r}
library(performance)

check_model(model)
```

```{r}
r2(model)
check_collinearity(model)
```

`check_model()` bundles several checks — linearity, homogeneity of variance, normality of residuals, influential observations, and collinearity — into one panel. `check_collinearity()` reports Variance Inflation Factors (VIF) on their own; values above roughly 5 suggest predictors are too correlated with each other to interpret their individual coefficients reliably.

:::
