---
title: "Statistical Tests and Analysis"
format:
  html:
    code-copy: true
    theme: cosmo
    toc: true
    lightbox: auto
---

::: {.content-visible when-profile="moderate"}

## Statistical Tests in R

This section covers common statistical tests used in medical research, implemented in R.

### Descriptive Statistics

```{r}
# Load necessary libraries
library(tidyverse)
library(gtsummary)

# Example data
data <- data.frame(
  age = rnorm(100, 50, 10),
  blood_pressure = rnorm(100, 120, 15),
  treatment = sample(c("A", "B"), 100, replace = TRUE)
)

# Summary statistics
data %>% 
  tbl_summary()
```

### Hypothesis Testing

#### T-test

```{r}
# Two-sample t-test
t.test(blood_pressure ~ treatment, data = data)
```

#### Chi-square test

```{r}
# Create categorical data
data$outcome <- sample(c("Improved", "Not Improved"), 100, replace = TRUE)

# Chi-square test
chisq.test(table(data$treatment, data$outcome))
```

### Regression Analysis

Regression modeling now has its own dedicated module — see **[Regression Analysis](regression.qmd)** for simple/multiple regression, diagnostics, and logistic regression.

:::